MFHD5: SC2

### 2018-03-29 mfhd5-sc2-ioslides.rmd
done sc2

---
title: "MFHD5: SC2"
author: "Aaron Khoo"
date:  `r format(Sys.Date(), '%d %B %Y')`
output:
  ioslides_presentation:
    incremental: true
    smaller: true
---

@@

-  **What is the current cervical screening program in Australia? What is its purpose? Where is the data stored?**
-  Current:
-  Purpose
-  Data: "securely" stored in the Register, protected by latest "state-of-the-art-data security" measures, according to "strict" Australian government info security requirements. Stored on-shore in Australia. Commonwealth Department of Health will have control over all info in the Register.
-  **Who is eligible/should be targeted for screening**
-  **What are some reasons for women not taking up cervical screening?**
-  **What are the risk factors for development of cervical cancer?**
-  HPV infection (HPV protein E7 cuts Rb brakes; E6 silences and degrades *TP53*), cigarette smoking (increases genetic damage via carcinogens), lack of regular screening
-  **How is the test reported, and what would be the process for an abnormal result?**

-  **What symptoms may patients with cervical pathology present with?**
-  **What are the barriers to health promotion in culturally and linguistically diverse (CALD) patients? How might you go about addressing these?**

@@

-  **How in practice is the cervical sample actually taken? How would you explain this to a patient?**
1.  After the patient is prepared and comfortable, begin the speculum examination by examining the external genitalia for any abnormalities, then:

@@

@@

-  **Why use an independent interpreter rather than her husband or other family member?**
-  Increased medical knowledge, better understanding of languages, covered under insurance, confidentiality
-  **Do you understand what is meant by the reference to being ‘cut’? What is female genital cutting / mutilation (FGM) (a term often best avoided), and what is its prevalence worldwide?** [more info](http://www.who.int/mediacentre/factsheets/fs241/en/)
-  Female genital mutilation (FGM) comprises all procedures that involve partial or total removal of the external female genitalia, or other injury to the female genital organs for non-medical reasons. It has no health benefits for girls and women. Procedures can cause severe bleeding and problems urinating, and later cysts, infections, as well as complications in childbirth and increased risk of newborn deaths.
More than 200 million girls and women alive today have been cut in 30 countries in Africa, the Middle East and Asia where FGM is concentrated. *It is a violation of  human rights*.
-  **Why might she have dyspareunia and post-coital bleeding?**
-  **What other complications are associated with the practice of FGM?**
-  Short term: pain, haemorrhage, shock, infn, HIV (possibly due to lack of sterlisation), urinary retention/dysuria, scarring, shock, psychological trauma
-  Long term: pain, infn (UTIs, chronica genital infns, chronic repro tract infns), dysuria, menstrual issues (due to obstruction of vaginal opening), keloids, HIV, female sexual Hx (removal of highly sensitive tissue, particularly clitoris, may decrease desire, pleasure, leading to dyspareunia, decreased lubrication, reduced frequency/absence of orgasm [anorgasmia], mental trauma), obstetric complications (such as caesarean, post-partum haemorrhage, tears/lacerations, prolonged labour), perinatal risks, psychological consequences
-  **What are the options if a speculum examination is too difficult in the GP practice?**
-   **What pathology of the cervix might be directly visualized on speculum examination?** [reading with some images](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3031439/)
-  Nabothian cysts: common & benign. May be translucent or opaque, whitish to yellow, mm to 4cm in diameter. Caused by transformation squamous metaplasia/inflamm blocking a gland orifice => endocervical columnar cells secrete mucus => covered by epithelium => cyst formation.
-  Leiomyomas
-  Cervical ectropion: eversion of the endocervix exposes columnar epithelium to vaginal milieu. Vaginal discharge most common symptom. Epithelium has reddish appearance sim to granulation tissue.
-  Cervical polyps: larger polyps should be evaluated and removed by a gynecologist; removal is typically straightforward office procedure
-  Endometriosis (present as red, blue, or black cervical lesions (“powder burns”)) and adenomyosis (may involve the endocervical canal or form a polypoid mass protruding into the endocervical canal)
-  Cervicitis: appearance depends on cause such as HSV (multiple small vesicular or ulcerative lesions), trichomonoiasis (punctate haemorrhages => strawberry cervix), chlamydial and gonococcal (mucopurulent discharge, cervical friability, cervical oedema)

@@

-  **Do you understand the terms ‘endocervix’, ‘transformation zone’, and ‘ectropion’?**
-  Transformation zone: between the endo and ectocervix which have two different epithelial linings. Endocervix. This is where HPV infects
-  **How does the transformation zone ‘move’ throughout reproductive life?**
-  Glandular epithelium moves down to ectocervix during female maturation
-  **What is colposcopy? Why and when is this necessary?**
-  **How would you explain the new cervical screening test to your patients who are used to having a Pap smear? What are the major changes you’d explain to them?**
-  Now it is every 5 years instead of 2 years (this is fine as the development is so slow), cheaper and safer, and pap cytology has been replaced by HPV molecular genotypinmg test
-  **Why do women who have had the HPV vaccine still need this screening? What is the theory behind the HPV cancer vaccine? When is it given in Australia?**
-  Gardasil targets 6/11/16/18, cervical cancer can also be caused by non-HPV 16/18 (25%) and non-HPV related (5%)

>>>>

---
title: "MFHD5: SC2"
author: "Aaron Khoo"
date:  "`r format(Sys.Date(), '%d %B %Y')`"
output:
  ioslides_presentation:
    incremental: true
    smaller: true
---

@@

-  [Use this reading for full info](http://www.cancerscreening.gov.au/internet/screening/publishing.nsf/Content/C2058A7D155867ACCA2581C400082790/$File/CAN174-Understanding-the-National-Cervical-Screening-Program-Management-Pathway.pdf)
-  **What is the current cervical screening program in Australia? What is its purpose? Where is the data stored?**
-  **How is the test reported, and what would be the process for an abnormal result?**
-  Current: occurs every 5 years. HPV testing uses partial genotyping to classify into oncogenic HPV 16/18 OR oncogenic HPV not-16/18. If HPV positive, then reflex LBC (liquid based cytology: placing the cervical sample in liquid suspension, rather than directly applying the sample on a microscope slide like a conventional Pap test) performed.
-  <img src="images/mfhd5-risk-cervical-ca.png" width="400" />

## Questions (cont.)

-  Purpose: to prevent cancer by detecting early and treating early
-  Data: "securely" stored in the Register, protected by latest "state-of-the-art-data security" measures, according to "strict" Australian government info security requirements. Stored on-shore in Australia. Commonwealth Department of Health will have control over all info in the Register.
-  **Who is eligible/should be targeted for screening**
  -  Anyone with a cervix (inc transgender men, intersex, females,...), 25 to 74 years, asymptomatic/symptomatic (e.g. abnormal vaginal bleeding, unusual vaginal discharge, dyspareunia)/pregnant
  -  Self collecting: 30 years or older, cannot be/think they are pregnant, declined to have cervical sample taken by clinician, and either 1) overdue by 2 or more years, 2) have never screened
-  **What are some reasons for women not taking up cervical screening?**
-  Don't want to, low access to health services, don't know about it
-  **What are the risk factors for development of cervical cancer?**
-  HPV infection (HPV protein E7 cuts Rb brakes; E6 silences and degrades *TP53*), cigarette smoking (increases genetic damage via carcinogens), lack of regular screening

## Questions (cont. 2)

-  **What symptoms may patients with cervical pathology present with?**
-  Vaginal bleeding, contact bleeding, loss of appetite, weight loss, fatigue, pelvic/back/leg pain, bone fractures
-  **What are the barriers to health promotion in culturally and linguistically diverse (CALD) patients? How might you go about addressing these?**
-  Language differences: refer to a clinician that speaks the language, use an interpreter, show the pamphlets in other language
-  Cultural differences: cultural sensitivity courses

@@

-  **How in practice is the cervical sample actually taken? How would you explain this to a patient?** [geekymedic is a great resource for OSCE stuff](https://geekymedics.com/cervical-smear-test/)
1.  After the patient is prepared and comfortable, begin the speculum examination by examining the external genitalia for any abnormalities
2.  Inspect cervix
3.  The objective of cervical screening is to sample cells from the transformation zone of the cervix, where HPV is present and cell abnormalities that precede the development of squamous cell carcinoma are usually found
4.  After, or before, the sample has been collected, record the patient name, DOB and ID number on the specimen vial and any other patient identifiers required by the laboratory, and fill out pathology request form
5.  After sample collection, ensure the details of the consultation and procedure are accurately documented in the patient’s clinical record

@@

## Trigger 2 (cont.)

@@

-  **Why use an independent interpreter rather than her husband or other family member?**
-  Increased medical terminology knowledge, better understanding of language, covered under insurance and other legal stuff, confidentiality
-  **Do you understand what is meant by the reference to being ‘cut’? What is female genital cutting / mutilation (FGM) (a term often best avoided), and what is its prevalence worldwide?** [more info](http://www.who.int/mediacentre/factsheets/fs241/en/)
-  Female genital mutilation (FGM) comprises all procedures that involve partial or total removal of the external female genitalia, or other injury to the female genital organs for non-medical reasons. It has no health benefits for girls and women. Procedures can cause severe bleeding and problems urinating, and later cysts, infections, as well as complications in childbirth and increased risk of newborn deaths.  
More than 200 million girls and women alive today have been cut in 30 countries in Africa, the Middle East and Asia where FGM is concentrated. *It is a violation of  human rights*.

## Questions (cont.)

-  **Why might she have dyspareunia and post-coital bleeding?**
-  **What other complications are associated with the practice of FGM?**
-  Short term: pain, haemorrhage, shock, infn, HIV (possibly due to lack of sterlisation), urinary retention/dysuria, scarring, shock, psychological trauma
-  Long term: pain, infn (UTIs, chronica genital infns, chronic repro tract infns), dysuria, menstrual issues (due to obstruction of vaginal opening), keloids, HIV, female sexual Hx (removal of highly sensitive tissue, particularly clitoris, may decrease desire, pleasure, leading to dyspareunia, decreased lubrication, reduced frequency/absence of orgasm [anorgasmia], mental trauma), obstetric complications (such as caesarean, post-partum haemorrhage, tears/lacerations, prolonged labour), perinatal risks, psychological consequences
-  **What are the options if a speculum examination is too difficult in the GP practice?**
-  Refer

## Questions (cont. 2)

-   **What pathology of the cervix might be directly visualized on speculum examination?** [reading with some images](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3031439/)
-  Nabothian cysts: common & benign. May be translucent or opaque, whitish to yellow, mm to 4cm in diameter. Caused by transformation squamous metaplasia/inflamm blocking a gland orifice => endocervical columnar cells secrete mucus => covered by epithelium => cyst formation.
-  Leiomyomas
-  Cervical ectropion: eversion of the endocervix exposes columnar epithelium to vaginal milieu. Vaginal discharge most common symptom. Epithelium has reddish appearance sim to granulation tissue.
-  Cervical polyps: larger polyps should be evaluated and removed by a gynecologist; removal is typically straightforward office procedure
-  Endometriosis (present as red, blue, or black cervical lesions (“powder burns”)) and adenomyosis (may involve the endocervical canal or form a polypoid mass protruding into the endocervical canal)
-  Cervicitis: appearance depends on cause such as HSV (multiple small vesicular or ulcerative lesions), trichomonoiasis (punctate haemorrhages => strawberry cervix), chlamydial and gonococcal (mucopurulent discharge, cervical friability, cervical oedema)

@@

-  **Do you understand the terms ‘endocervix’, ‘transformation zone’, and ‘ectropion’?** [this site more info](https://www.gfmer.ch/ccdc/pdf/module1.pdf)
-  **How does the transformation zone ‘move’ throughout reproductive life?**
-  Cervix = ectocervix (visible during speculum exam) + endocervix
-  Endocervix: luminal cavity within the cervix forming a passageway between the external os and the internal os. Lined by simple columnar cells that secrete mucous
-  Ectocervix lined by simple non-keratinizing stratified squamous epithelium
-  The squamocolumnar junction (SCJ): junction between the squamous epithelium and the columnar epithelium. Located at the external os pre-puberty, repro age is variable, post menopausal the SCJ is not visible and has receded into the endocervix.
-  Ectropion: eversion of squamocolumnar junction and large portions of columnar epithelium onto the ectocervix. This is not an abnormality and it can be caused by things such as hormonal changes, taking the OCP, pregnancy. However it is indistinguishable from early cervical cancer. When the everted columnar epithelium is exposed to acidic vaginal environment => squamous metaplasia => stratified squamous
-  Transformation zone (TZ): cervical area between the old SCJ and new SCJ. >90% of precancerous lesions begin here

## Questions (cont.)

-  **What is colposcopy? Why and when is this necessary?**
-  The cervix is inspected and acetic acid is applied to its surface – abnormal cells have increased surface proteins, which are coagulated by the acid and turn white, producing “acetowhite” areas which can be biopsied for histological assessment. Necessary: see above table
-  **How would you explain the new cervical screening test to your patients who are used to having a Pap smear? What are the major changes you’d explain to them?**
-  Now it is every 5 years instead of 2 years (this is fine as the development is so slow), cheaper and safer, and pap cytology has been replaced by HPV molecular genotyping test
-  **Why do women who have had the HPV vaccine still need this screening? What is the theory behind the HPV cancer vaccine? When is it given in Australia?**
-  Gardasil targets 16/18, cervical cancer can also be caused by non-HPV 16/18 (25%) and non-HPV related (5%)
